---
title: "Local food webs and their indices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Local food webs and their indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 90
)
```

```{r load}
library(metawebr)
library(ggplot2)
```

## Overview

A metaweb lists every feeding link that could exist among the taxa of a
region. A *local* food web is the part of the metaweb formed by the taxa found
together at one site, for example in an eDNA sample or a survey: a link is kept
when both taxa occur at the site. This vignette builds local food webs,
computes their network indices, compares them with null models and plots them.

| Step | What happens | Function(s) |
|------|--------------|-------------|
| 1 | Match the site data to the metaweb | `rename_taxa()` |
| 2 | Build the site × taxon matrix | `make_species_station_matrix()` |
| 3 | Compute local indices | `get_indic_cells()` |
| 4 | Compare indices with null models | `get_indic_null()`, `plot_indic_null()` |
| 5 | Plot local food webs | `plot_tree_network()`, `plot_network_focus_sp()`, `plot_network_interactive()` |

It starts from a binary metaweb. Building and curating one is covered in
`vignette("metaweb", package = "metawebr")`; the metaweb built there is shipped
as `metaweb_obis`:

```{r metaweb}
mw_bin <- metaweb_obis
dim(mw_bin)
sum(mw_bin)  # number of links
mw_bin[c("Chromis_chromis", "Coris_julis"), c("Muraena_helena", "Serranus_scriba")]
```

As everywhere in the package, **rows are prey and columns are predators**:
`mw_bin["Coris_julis", "Muraena_helena"] == 1` means that the moray eel eats
the wrasse.

## 1. Site data and taxon names

`df_presence_obis` lists the fish recorded at six sites from OBIS data: three
Mediterranean rocky reefs (Marseille, Medes Islands, Port-Cros) and three
tropical western Atlantic coral reefs (Florida Keys, Belize Barrier Reef,
Bonaire). It is in long format, one row per site and species, with a count.

```{r presence}
head(df_presence_obis)
table(df_presence_obis$station)
```

Local webs can only contain taxa that are in the metaweb, under the same name.
Check which detected taxa are missing:

```{r missing}
setdiff(unique(df_presence_obis$species), rownames(mw_bin))
```

*Equetus punctatus* is not in FishBase and was left out of the metaweb. With
your own data, missing names are often synonyms or spelling variants: harmonise
the site data with the same names as the metaweb before going further:

```{r harmonise, eval = FALSE}
harm <- harmonise_taxa(unique(detections$taxon))
detections <- rename_taxa(detections, "taxon", harm)
```

## 2. Site × taxon matrix

`make_species_station_matrix()` turns long detection data (one row per site ×
taxon, with a count such as eDNA reads or records) into a binary site × taxon
matrix. Taxa detected several times at a site, or several names merged by
`rename_taxa()`, are summed. It also adds the two producer nodes
(`PrimaryProducer`, `SecondaryProducer`) to every site, since primary
production and invertebrates are assumed present everywhere.

```{r stations}
pa <- make_species_station_matrix(df_presence_obis,
                                  station_col = "station",
                                  species_col = "species",
                                  count_col = "count")
dim(pa)
pa[, c("Coris_julis", "Thalassoma_bifasciatum", "PrimaryProducer")]
```

`get_indic_cells()` drops taxa that are absent from the metaweb, with a
warning. We drop *Equetus punctatus* explicitly:

```{r stations-filter}
pa <- pa[, colnames(pa) %in% rownames(mw_bin)]
```

## 3. Local indices

`get_indic_cells()` extracts the local web of each site and returns one row per
site with about 45 indices. Its `mc.cores` argument runs sites in parallel (not
supported on Windows).

```{r indices}
indices <- get_indic_cells(P_A_data = pa, Lniche = mw_bin)

cols <- c("Species_taxa", "Link_taxa", "Connectance_taxa", "TL_moy", "length_chain",
          "Omn_moy", "Nbas", "Nint", "Ntop", "Modularity")
cbind(site = indices$site, round(indices[, cols], 3))
```

The indices fall into four groups:

| Group | Index | Meaning |
|-------|-------|---------|
| Size and links | `Species`, `Link`, `Connectance` | Number of nodes (including the two producer nodes), number of links, and links / nodes² |
| | `Species_taxa`, `Link_taxa`, `Connectance_taxa` | The same without the producer nodes |
| | `Connectance_p`, `b_power_law` | Connectance from link values, and the exponent of the link–species scaling |
| Trophic structure | `TL_moy`, `length_chain` | Mean trophic level, and maximum trophic level rounded up |
| | `Omn_moy` | Mean omnivory index (spread of the trophic levels of the prey) |
| | `Nbas`, `Nint`, `Ntop` | Fractions of basal (no prey among the taxa: eat only producers), intermediate, and top (not basal, no predator) taxa; they sum to 1 |
| | `Planktivores`, `HTI`, `MTI` | Share of taxa with a trophic level in (2.4, 3.7), share above 4, and mean trophic level of taxa above 3.25, producers excluded. Only for webs with more than `min_species_tl` taxa (default 50), otherwise `NA` |
| Links per taxon | `Vul`, `Vulsd`, `Gen`, `Gensd` | Mean and sd of the number of predators (vulnerability) and of prey (generality) |
| | `Din_*`, `Dout_*` | Mean, min and max in- and out-degree |
| Network shape | `Modularity` | Modularity of walktrap communities (`igraph`) |
| | `Mean_path_length`, `Shortest_path_k`, `Diameter` | Mean shortest path, share of shortest paths of length *k*, longest shortest path |
| | `Transitivity`, `Closeness_*`, `Coreness_*`, `Nb_articulate_point` | Clustering, closeness, k-core and articulation-point statistics |

See `?get_indic_cells` for the exact definitions.

The sites here hold 40 species each, below the default `min_species_tl = 50`,
so `Planktivores`, `HTI` and `MTI` are `NA`. These shares are unstable in small
webs; lower the cut-off only if you accept that:

```{r indices-tl}
indices_tl <- get_indic_cells(pa, mw_bin, min_species_tl = 30)
cbind(site = indices_tl$site, round(indices_tl[, c("Planktivores", "HTI", "MTI")], 3))
```

Comparing the two regions:

```{r indices-plot, fig.height = 4}
region <- setNames(df_presence_obis$region, df_presence_obis$station)
long <- data.frame(
  site = indices$site,
  region = unname(region[indices$site]),
  index = rep(c("Connectance_taxa", "TL_moy", "Omn_moy", "Modularity"), each = nrow(indices)),
  value = unlist(indices[, c("Connectance_taxa", "TL_moy", "Omn_moy", "Modularity")], use.names = FALSE)
)
ggplot(long, aes(region, value, colour = region)) +
  geom_point(size = 3) +
  facet_wrap(~index, scales = "free_y", nrow = 1) +
  labs(x = NULL, y = NULL) +
  theme_bw() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 30, hjust = 1))
```

Raw indices are hard to compare between sites with different numbers of
species, because many of them change with richness. Null models address this.

## 4. Null models

Many indices change with the number of species, so raw values are hard to
compare between sites. A null model asks whether a local web is more or less
connected, modular, etc. than expected for a random community of the same
richness, drawn from the regional pool.

`get_indic_null()` builds `n_null` random communities per site, keeping the
site's richness and its producer nodes, and computes the same indices. For each
index it reports:

* the null mean and standard deviation;
* the standardised effect size, `ses = (observed − null_mean) / null_sd`.
  Values below −1.96 or above 1.96 are roughly significant at 5 %;
* a two-sided, rank-based `p_value`.

Two methods are available:

* `"equiprobable"`: every taxon of the pool is equally likely to be drawn;
* `"frequency"`: taxa are drawn in proportion to the number of sites where they
  occur, so common taxa appear more often.

The pool is every taxon present in at least one site of `P_A_data`. Pooling the
Mediterranean and Caribbean sites would put Caribbean species into
Mediterranean null communities, so we run the null model separately for each
region:

```{r null-equiprobable}
sites_by_region <- split(rownames(pa), region[rownames(pa)])

set.seed(42)
null_equi <- do.call(rbind, lapply(sites_by_region, function(sites) {
  get_indic_null(pa[sites, ], mw_bin, method = "equiprobable", n_null = 199)
}))
rownames(null_equi) <- NULL

res <- null_equi[null_equi$index %in% c("Connectance", "TL_moy"), ]
cbind(res[, c("site", "index")], round(res[, c("observed", "null_mean", "ses", "p_value")], 3))
```

`plot_indic_null()` shows the effect sizes of several indices at once. Points
are coloured when p < 0.05 and dashed lines mark SES = ±1.96:

```{r null-plot, fig.width = 8, fig.height = 5.5}
plot_indic_null(null_equi,
                indices = c("Connectance", "TL_moy", "Omn_moy",
                            "Modularity", "Mean_path_length", "Nbas"))
```

`get_indic_cells()` can compute the observed indices and the null model in one
call. With `null_model = "none"` (the default) it returns the indices only;
with a method it returns a list with `indices` and `null_model`. Here is the
frequency-weighted model for the Mediterranean sites, compared with the
equiprobable one:

```{r null-frequency}
set.seed(42)
med <- get_indic_cells(pa[sites_by_region$Mediterranean, ], mw_bin,
                       null_model = "frequency", n_null = 199)
names(med)

compare <- merge(null_equi[, c("site", "index", "ses")],
                 med$null_model[, c("site", "index", "ses")],
                 by = c("site", "index"), suffixes = c("_equiprobable", "_frequency"))
compare[, c("ses_equiprobable", "ses_frequency")] <-
  round(compare[, c("ses_equiprobable", "ses_frequency")], 2)
compare[compare$index %in% c("Connectance", "TL_moy"), ]
```

`n_null = 199` keeps this vignette fast. Use 999 or more for published analyses.

## 5. Plotting food webs

`plot_tree_network()` draws a food web with each node's height set by its
trophic level. Nodes are coloured by rounded trophic level and sized by number
of prey. Horizontal positions are random, so set a seed for a reproducible
layout. The result is a `ggplot`, so you can add titles and themes as usual.
When nodes are crowded, some labels are left out.

Use `sp_to_keep` to plot a local web. Trophic levels are still computed from the
full metaweb, so node heights are comparable between sites:

```{r plot-local, fig.width = 9, fig.height = 7, warning = FALSE}
marseille <- df_presence_obis$species[df_presence_obis$station == "Marseille"]
set.seed(1)
plot_tree_network(mw_bin, sp_to_keep = marseille) + ggtitle("Marseille")
```

`plot_network_focus_sp()` zooms in on one or more focal taxa and their prey, up
to `order_interaction` steps down the food chain (`2` adds the prey of the prey):

```{r plot-focus, fig.width = 8, fig.height = 6, warning = FALSE}
set.seed(1)
plot_network_focus_sp(mw_bin, sp_focus = "Muraena_helena",
                      order_interaction = 2, sp_to_keep = marseille) +
  ggtitle("Moray eel and its prey in Marseille")
```

`plot_network_interactive()` draws the same layout as an interactive network
(it needs the visNetwork package). Click a taxon to highlight its links and
fade the rest of the web; hold Ctrl (Cmd on macOS) while clicking to select
several taxa, or pick one by name in the menu. Click an empty area to clear
the selection. Hover a taxon or a link for details, scroll to zoom and drag to
pan. With `links = "prey"` or `links = "predators"` only one direction is
highlighted:

```{r plot-interactive, eval = requireNamespace("visNetwork", quietly = TRUE)}
set.seed(1)
plot_network_interactive(mw_bin, sp_to_keep = marseille, height = "600px")
```

The result is an htmlwidget, so you can save it as a stand-alone HTML file to
share:

```{r save-interactive, eval = FALSE}
web <- plot_network_interactive(mw_bin, sp_to_keep = marseille)
htmlwidgets::saveWidget(web, "marseille_foodweb.html")
```

## Good practice

* **The metaweb gives potential links.** A local web keeps every link of the
  metaweb between co-occurring taxa, whether or not it happens at that site.
  Indices describe the potential structure of the community.
* **Detection matters.** Taxa missed by the survey or the eDNA assay, and
  taxa identified only to genus or family, change the local web. Report how
  taxa were identified and harmonised.
* **Producers are present everywhere.** The two producer nodes are added to
  every site. Prefer the `*_taxa` indices when comparing sites with very
  different numbers of fish.
* **Compare with null models.** Use them before interpreting differences
  between sites, with a regional pool that makes sense (here, one pool per
  region), and 999 or more null communities for published analyses.
* **Check sensitivity to the threshold.** Rebuild the binary metaweb at a few
  thresholds around the chosen one (`correct_metaweb_proba(threshold = )`) and
  check that your conclusions hold.

## References

* Gravel, D., Poisot, T., Albouy, C., Velez, L. & Mouillot, D. (2013). Inferring
  food web structure from predator–prey body size relationships.
  *Methods in Ecology and Evolution*, 4, 1083–1090.
* Albouy, C. et al. (2019). The marine fish food web is globally connected.
  *Nature Ecology & Evolution*, 3, 1153–1161.
* Froese, R. & Pauly, D. (eds.). FishBase. World Wide Web electronic publication, www.fishbase.org.
* Rabosky, D.L. et al. (2018). An inverse latitudinal gradient in speciation
  rate for marine fishes. *Nature*, 559, 392–395.
* Stekhoven, D.J. & Bühlmann, P. (2012). MissForest: non-parametric missing
  value imputation for mixed-type data. *Bioinformatics*, 28, 112–118.
* OBIS. Ocean Biodiversity Information System. Intergovernmental Oceanographic
  Commission of UNESCO. www.obis.org.
