---
title: "Building and curating a fish metaweb"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building and curating a fish metaweb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 90
)
```

```{r load}
library(metawebr)
library(ggplot2)
```

## Overview

A *metaweb* is the regional food web of all the taxa that could occur
together: who could eat whom, whether or not they meet at a given site.
`metawebr` infers it from body size, curates it with ecological rules and
diet data, and checks it against observed feeding links. This vignette covers
these steps. The companion vignette,
`vignette("local-food-webs", package = "metawebr")`, extracts local food webs
from the metaweb and computes their indices.

| Step | What happens | Function(s) |
|------|--------------|-------------|
| 1 | Explore predator–prey body-size records | `df_interaction_fish` |
| 2 | Calibrate the body-size niche model | `metaweb_mod_parameters()`, `get_fig_eval()` |
| 3 | Harmonise taxon names | `harmonise_taxa()`, `rename_taxa()` |
| 4 | Gather and clean traits | `get_traits()`, `infer_traits()`, `clean_traits()`, `impute_traits_phylo()` |
| 5 | Predict interaction probabilities for every pair of taxa | `apply_model_metaweb()` |
| 6 | Curate the metaweb with ecological rules and diet data | `set_herbivores()`, `add_depth_range()`, `correct_metaweb_fish()`, `compare_metawebs()` |
| 7 | Choose a probability threshold and binarise | `get_proba()`, `correct_metaweb_proba()` |
| 8 | Validate the size model and the metaweb | `cv_size_model()`, `validate_metaweb()`, `get_diet_links()` |

The interaction model is the allometric niche model of Gravel et al. (2013).
A predator of body length $L_{pred}$ feeds on prey whose length falls within a
Gaussian niche. The niche optimum $o$ and width $r$ both scale with predator
size (all lengths on a log10 scale):

$$
o = a_0 + a_1 \log_{10} L_{pred}, \qquad
r = b_0 + b_1 \log_{10} L_{pred}, \qquad
P(\text{link}) = \exp\left(-\frac{(o - \log_{10} L_{prey})^2}{2r^2}\right)
$$

FishBase data (traits, synonyms, diets, depth ranges) are downloaded once per
R session. To keep them between sessions, set a cache folder:
`options(metawebr.cache_dir = "~/fishbase_cache")` (see
`?clear_fishbase_cache`). Steps that need an internet connection are shown but
not run in this vignette; their output is pasted as comments.

## 1. Predator–prey body-size data

The package ships `df_interaction_fish`, a compilation of `r nrow(df_interaction_fish)`
observed feeding events. Each row gives a predator, a prey and their lengths (cm).
Only four columns are used for calibration:

```{r data}
data("df_interaction_fish")
head(df_interaction_fish[, c("predator", "prey",
                             "standardised_predator_length", "si_prey_length")])
```

On a log scale, prey size increases with predator size, and the spread of prey
sizes widens too. The niche model captures exactly this pattern:

```{r data-plot, fig.height = 4.5}
ggplot(df_interaction_fish,
       aes(standardised_predator_length, si_prey_length)) +
  geom_point(alpha = 0.15, size = 0.8) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Predator length (cm, log scale)",
       y = "Prey length (cm, log scale)") +
  theme_minimal()
```

## 2. Calibrate the niche model

`metaweb_mod_parameters()` removes duplicate records and caps each
predator–prey pair at `observation_resample` observations, so that heavily
sampled pairs don't dominate the fit. It then estimates $a_0, a_1, b_0, b_1$ by
maximum likelihood, using simulated annealing (`GenSA`). With
`data_path = NULL` it uses `df_interaction_fish`; you can also pass a data
frame or the path of a CSV with the same columns.

Calibration takes about two minutes and involves random down-sampling, so it is
not run here:

```{r calibrate, eval = FALSE}
set.seed(1)
params <- metaweb_mod_parameters(path_par_output = "params_metaweb.txt")
params$calibrated_data
```

The other functions accept the parameters in three forms: the list returned
by `metaweb_mod_parameters()`, the file it wrote, or a plain numeric vector.
This vignette uses the values estimated from `df_interaction_fish`:

```{r pars}
pars <- c(a0 = -0.5616698, a1 = 0.9611534, b0 = 0.1247726, b1 = 0.1359744)
```

`get_fig_eval()` checks the fit for one predator. The left panel compares the
sizes of that predator's observed prey with all prey sizes and with the
predicted niche. The right panel shows the predicted link probability across
the whole predator–prey size space, overlaid with the observations.

```{r fit-eval, fig.width = 9, fig.height = 4.5}
get_fig_eval(data = df_interaction_fish, target_sp = "Gadus morhua", pars = pars)
```

## 3. Example taxa and their names

The metaweb is built for a list of taxa: here, the species of
`df_presence_obis`, an example community dataset built from OBIS occurrence
records since 2000. It holds the 40 most recorded fish species at each of six
sites:

* three Mediterranean rocky reefs: Marseille, Medes Islands, Port-Cros;
* three tropical western Atlantic coral reefs: Florida Keys, Belize Barrier
  Reef, Bonaire.

```{r presence}
data("df_presence_obis")
head(df_presence_obis)
table(df_presence_obis$region)
```

### Harmonising taxon names

Names from eDNA reference databases (often NCBI taxonomy) or surveys don't
always match FishBase: synonyms (*Liza aurata* is now *Chelon auratus*),
spelling mistakes, `Genus sp.` assignments, subspecies. Unmatched names get no
traits and are silently lost, so harmonise them first. `harmonise_taxa()`
cleans each name, then looks it up in your own correction table (`lookup`),
then in FishBase (accepted names and synonyms), and finally in WoRMS for the
names still unresolved. The accepted FishBase name is the standard used by the
rest of the package; the WoRMS AphiaID is kept for linking to other data.

```{r harmonise, eval = FALSE}
harm <- harmonise_taxa(c("Liza aurata", "Diplodus_sp.", "Gadus morua",
                         "Pomatoschistus minutus/lozanoi", "uncultured fish"))
harm[, c("input", "accepted", "status", "note")]
#>                            input       accepted       status                        note
#> 1                    Liza aurata Chelon_auratus      synonym                        <NA>
#> 2                   Diplodus_sp.       Diplodus     accepted          sp. dropped: genus
#> 3                    Gadus morua   Gadus_morhua  worms_fuzzy                        <NA>
#> 4 Pomatoschistus minutus/lozanoi Pomatoschistus     accepted several species: genus kept
#> 5                uncultured fish           <NA>  not_a_taxon                        <NA>

# Replace the names in the detection table, keeping the originals
detections <- rename_taxa(detections, "taxon", harm)
```

Check the names with status `worms_fuzzy` (spelling corrected), `ambiguous`
and `unresolved`, and record your decisions in a `lookup` table
(`data.frame(input = , accepted = )`) so the harmonisation is reproducible.

## 4. Traits

The model needs one row per taxon, with **row names set to the taxon names**,
and these columns:

* `CommonLengthEstim`: common length (cm). If FishBase has no common length,
  it is estimated as 0.6 × maximum length.
* `TrophicLevel`: the FishBase trophic level estimate. It is used by the
  corrections and to choose the threshold.
* `DemersPelag`: position in the water column (FishBase categories).

`get_traits()` downloads these traits from FishBase through `rfishbase`.
Taxon names must use underscores (`Coris_julis`). For eDNA taxa assigned only
to a genus or a family, `infer_traits()` averages the traits of all FishBase
species in that group.

```{r traits-fishbase, eval = FALSE}
df_traits_obis <- get_traits(df_presence_obis, column_species = "species")
```

The result is stored as `df_traits_obis`, so this vignette runs offline.
`clean_traits()` reports missing values and fills them from other species of
the same genus. By default it then removes taxa that still have missing traits.
Here *Equetus punctatus* is not in FishBase and has no congener in the data, so
it is removed:

```{r traits}
data("df_traits_obis")
traits <- clean_traits(df_traits_obis)
```

`clean_traits()` only borrows values from congeners in your own table. A more
thorough option is `impute_traits_phylo()`. It predicts missing values with
random forests (missForest), using each species' position in the fish
phylogeny as extra information. It repeats this over the 100 complete fish
trees of Rabosky et al. (2018) and averages the results. It downloads the trees
and takes a few minutes, so it is not run here:

```{r traits-impute, eval = FALSE}
set.seed(1)
traits_imputed <- impute_traits_phylo(df_traits_obis, mc.cores = 4)
traits_imputed[!is.na(traits_imputed$imputed), ]

# Or in one step from taxon names: FishBase lookup, then imputation
traits_obis <- infer_traits(unique(df_presence_obis$species), method = "100matrix",
                            mc.cores = 4)
```

The fishtree trees cover ray-finned fishes only. To impute sharks and rays,
pass your own trees with `trees =`.

## 5. Predict interaction probabilities

`apply_model_metaweb()` applies the calibrated model to every pair of taxa.
In the resulting square matrix, **rows are prey and columns are predators**:
`mw["Chromis_chromis", "Muraena_helena"]` is the probability that the moray
eel eats the damselfish.

```{r metaweb}
mw <- apply_model_metaweb(traits, pars)
dim(mw)
round(mw[c("Chromis_chromis", "Coris_julis", "Diplodus_vulgaris"),
         c("Muraena_helena", "Scorpaena_scrofa", "Coris_julis")], 2)
```

## 6. Curating the metaweb

Size alone predicts some implausible links: a large herbivore "eating" small
fish, or a deep-sea fish eating a reef fish. `correct_metaweb_fish()` applies
six corrections, each of which can be switched off:

* `herbivory`: herbivores have no fish prey. By default these are taxa with a
  trophic level < 2.4; on these reefs, mostly parrotfishes, surgeonfishes and
  damselfishes. They can also be set from FishBase diet data (see below).
* `water_pos`: removes links between taxa that live in incompatible parts of the
  water column (for example bathydemersal vs. pelagic-oceanic). It also removes
  cannibalism.
* `depth_overlap` (off by default): removes links between taxa whose depth
  ranges don't overlap.
* `small_fish`: taxa shorter than 10 cm have no fish prey.
* `PP_SP_add`: adds two basal nodes, `PrimaryProducer` and `SecondaryProducer`,
  as food for herbivores, planktivores and small fish.
* `exception_BigFish`: large (> 100 cm) low-trophic-level taxa feed on
  secondary producers only.

```{r correct}
mw_corr <- correct_metaweb_fish(data_meta = mw, data_traits = traits)
dim(mw_corr)
```

Every link a correction changes is recorded, with the correction and the data
it was based on (`source`), so you can see why each link was changed (see
*Checking the corrections* below).

### Herbivores

The first message lists the taxa treated as herbivores. FishBase trophic levels
can misclassify some taxa, so check this list. To change it, use
`set_herbivores()`. It stores the classification in a `Herbivore` column of the
trait table, which `correct_metaweb_fish()` then uses instead of the trophic
level. For example, to stop treating the scrawled cowfish as a herbivore:

```{r herbivores}
traits_edited <- set_herbivores(traits, remove = "Acanthostracion_polygonium")
sum(traits_edited$Herbivore)
```

You would then run `correct_metaweb_fish(mw, traits_edited)` again.

The trophic level is an indirect criterion: FishBase estimates it partly from
diet, and a cut-off at 2.4 misclassifies omnivores and detritivores.
`set_herbivores(source = "fishbase")` uses FishBase diet data instead: the
share of plant and detritus food items when a species has at least three
records, otherwise its main feeding type (e.g. grazing or browsing on
substrate). Taxa without diet data fall back on the trophic level. The
`Herbivore_source` column records which criterion was used for each taxon:

```{r herbivores-diet, eval = FALSE}
traits_diet <- set_herbivores(traits, source = "fishbase")
#> 19 herbivores (diet_items: 5, feeding_type: 13, trophic_level: 1).
traits_diet[xor(traits$TrophicLevel < 2.4, traits_diet$Herbivore),
            c("TrophicLevel", "Herbivore", "Herbivore_source")]
#>                             TrophicLevel Herbivore Herbivore_source
#> Chelon_labrosus                     2.59      TRUE       diet_items
#> Coryphopterus_glaucofraenum         2.67      TRUE     feeding_type
#> Acanthostracion_polygonium          2.03     FALSE       diet_items
```

The rest of this vignette keeps the default classification.

### Depth

Species of the same water-column category can still live hundreds of metres
apart. Depth ranges come from FishBase (`add_depth_range()`) or from the
`df_depth_env` table shipped with the package. Links between taxa that never
meet are then removed; `depth_tolerance` allows a gap in metres:

```{r depth}
traits_depth <- add_depth_range(traits, source = "df_depth_env")
mw_depth <- suppressMessages(suppressWarnings(
  correct_metaweb_fish(mw, traits_depth, depth_overlap = TRUE)
))
sum(attr(mw_depth, "corrections")$correction == "depth")
```

Depth ranges are those of the whole species, across life stages and regions,
so only clearly separated species lose their links. The rest of this vignette
keeps the default corrections.

### Checking the corrections

`compare_metawebs()` lists what the corrections changed, so you can check it by
hand. Each species pair appears once, with both directions of the interaction:

* `taxon_1`, `taxon_2`: the two taxa;
* `t1_eats_t2_before`, `t1_eats_t2_after`, `t1_eats_t2_correction`,
  `t1_eats_t2_source`: the probability that taxon 1 eats taxon 2 before and
  after correction, the correction that changed it (`NA` if unchanged) and the
  data it was based on (e.g. `trophic_level` or `diet_items` for herbivory,
  `DemersPelag` for water position);
* the same four `t2_eats_t1_*` columns for the other direction (`NA` for
  cannibalism, where both taxa are the same);
* the traits of both taxa (`t1_*`, `t2_*`).

Pairs are sorted by the highest probability, before correction, among their
changed links, so the most likely links that were removed come first. Changed
links below 0.01 both before and after are hidden (`min_proba = 0` shows them
all).

```{r corrections-pairs}
changes <- compare_metawebs(mw, mw_corr, traits)
nrow(changes)
# number of changed links per correction and source
counts <- as.data.frame(table(
  correction = c(changes$t1_eats_t2_correction, changes$t2_eats_t1_correction),
  source = c(changes$t1_eats_t2_source, changes$t2_eats_t1_source)
))
counts[counts$Freq > 0, ]

shown <- head(changes[, 1:10])
shown[, c(3, 4, 7, 8)] <- signif(shown[, c(3, 4, 7, 8)], 2)
shown
```

Use `level = "link"` instead for one row per direction.

With `level = "taxon"` you get one row per taxon, with how much each
correction changed its prey. For a probability metaweb these are expected
numbers of prey:

```{r corrections-taxa}
by_taxon <- compare_metawebs(mw, mw_corr, traits, level = "taxon")
cols <- c("prey_before", "prey_after", grep("^prey_change_", names(by_taxon), value = TRUE))
cbind(taxon = by_taxon$taxon, round(by_taxon[, cols], 1))[
  by_taxon$taxon %in% c("Muraena_helena", "Sarpa_salpa", "Tripterygion_delaisi"), ]
```

To review the changes in a spreadsheet, save the table with
`write.csv(changes, "metaweb_corrections.csv", row.names = FALSE)`.
`compare_metawebs()` works on any two metawebs, e.g. before and after
binarisation (`compare_metawebs(mw_corr, mw_bin)`); changes that were not made
by `correct_metaweb_fish()` have no `correction`.

## 7. Choose a threshold and binarise

Network indices need presence/absence links, so the probabilities must be
thresholded. `get_proba()` tries thresholds from 0.1 to 0.995. For each one it
computes the trophic levels implied by the binary web and sums their absolute
differences from the FishBase trophic levels. It keeps the threshold with the
smallest error:

```{r threshold, fig.height = 4.5}
threshold <- get_proba(mw_corr, traits)
threshold
```

`correct_metaweb_proba()` runs the same search and returns the binary metaweb.
You can also give it a `threshold` directly, for example to test how sensitive
your results are to this choice.

```{r binarise, fig.show = "hide"}
mw_bin <- correct_metaweb_proba(mw_corr, traits)
sum(mw_bin)  # number of links
```

The trophic-level criterion only constrains where taxa sit in the web on
average, not which links exist. When observed links are available,
`get_proba(method = "tss")` instead picks the threshold that best separates
observed links from other pairs (maximum true skill statistic, see below).

## 8. Validating the metaweb

How well does the metaweb predict who eats whom? Two complementary checks:

**1. Cross-validation of the size model.** The interaction data used for
calibration can't validate it directly. `cv_size_model()` leaves out one
study at a time (the `reference` column), calibrates the model on the other
studies and predicts the left-out one. Leaving out whole studies, rather than
random records, keeps near-identical records of the same study out of the
calibration.

Each predator of the left-out study is paired with every prey species of the
data (`prey_pool = "all"`). Pairs never recorded count as non-links. These are
pseudo-absences, so the scores are conservative. Most studies of
`df_interaction_fish` describe the diet of a single predator, so the stricter
`prey_pool = "study"` (only prey recorded in the left-out study) leaves them
with no non-links to score. The in-sample AUC (model calibrated on all
studies) shows how much the calibration overfits. Calibration runs once per
study, so this takes a few minutes:

```{r cv, eval = FALSE}
set.seed(1)
cv <- cv_size_model(max_time = 30)
cv$folds[, c("group", "n_predators", "n_links", "AUC", "AUC_in_sample", "TSS")]
#>                        group n_predators n_links  AUC AUC_in_sample   TSS
#> 1         Chase, B.C. (2002)           1      11 0.89          0.90 0.667
#> 2       Pinnegar et al. 2003          28     178 0.57          0.69 0.199
#> 3      Pusineri et al (2005)           1       6 0.49          0.49 0.096
#> 4    Chancollon et al (2006)           1      13 0.58          0.58 0.297
#> 5 Barraclough (1967) Rep 922           7       8 0.72          0.76 0.331
#> 6  Stergiou & Fourtouni 1991           1      21 0.64          0.64 0.292
#> 7       Kellermann, A. 1987.           1       0   NA            NA    NA
#> 8       Kellermann, A. 1989.           1       4   NA            NA    NA
cv$overall
#>           model   AUC   TSS threshold
#> 1 out_of_sample 0.626 0.263     0.150
#> 2     in_sample 0.706 0.337     0.375
```

An AUC of 0.5 is no better than chance and 1 is perfect. Body size alone
separates a predator's prey from other prey only moderately, and less well
for studies it was not calibrated on. This is why the ecological corrections
and diet data matter.

**2. Validation of the corrected metaweb** against independent links, with
`validate_metaweb()`. Links can come from your own stomach-content data, GloBI,
or FishBase diet records (`get_diet_links()`). It reports AUC, sensitivity,
specificity and TSS, and, for each predator, the observed prey the metaweb
misses:

```{r validate, eval = FALSE}
links <- get_diet_links(rownames(mw_corr))
#> 49 links for 16 predators found in FishBase diet records.
val <- validate_metaweb(mw_corr, links)
val$summary[, c("n_predators", "n_links", "AUC", "sensitivity", "specificity", "TSS")]
#>   n_predators n_links  AUC sensitivity specificity  TSS
#> 1          15      48 0.56        0.44         0.7 0.14
head(val$per_predator[, c("predator", "n_observed", "n_hits", "missed")], 3)
#>                  predator n_observed n_hits                                                                  missed
#> 1        Scorpaena_porcus          4      0 Chromis_chromis, Symphodus_tinca, Mullus_surmuletus, Symphodus_roissali
#> 2        Scorpaena_scrofa          4      0     Serranus_scriba, Diplodus_puntazzo, Diplodus_annularis, Boops_boops
#> 3 Cephalopholis_cruentata          3      0     Abudefduf_saxatilis, Coryphopterus_personatus, Azurina_multilineata

# Threshold chosen from observed links instead of trophic levels
mw_bin_tss <- correct_metaweb_proba(mw_corr, method = "tss", links = links)
```

Here the corrected metaweb recovers fewer than half of the FishBase links.
`per_predator` shows where it fails: scorpionfishes and groupers. The
corrections play no part; the size model itself gives these links low
probabilities, because the missed prey have adult lengths close to or above
the predator's (e.g. a 15 cm *Scorpaena porcus* eating 25 cm *Mullus
surmuletus*). The predators most likely eat juveniles, which a single adult
length per species can't represent. FishBase lists species-level prey for
few fish, so this check covers only 15 of the 135 predators here: use your own
diet data when you have them. Use
data that were not used to build the metaweb: if herbivores were set from
FishBase diet data, validating against FishBase diet links is not independent,
and `validate_metaweb()` warns about it. Likewise, a threshold chosen with
`method = "tss"` should be validated on other links.

## 9. Saving the metaweb

The binary metaweb is the input of the local analyses. Save it, together with
the traits and the parameters, so the local analyses use exactly this
version:

```{r save, eval = FALSE}
saveRDS(list(metaweb = mw_bin, traits = traits, pars = pars), "metaweb_obis.rds")
```

The metaweb built here is shipped with the package as `metaweb_obis`, which
the local food-web vignette uses:

```{r shipped}
all.equal(mw_bin, metaweb_obis, check.attributes = FALSE)
```

Continue with `vignette("local-food-webs", package = "metawebr")`.

## References

* Gravel, D., Poisot, T., Albouy, C., Velez, L. & Mouillot, D. (2013). Inferring
  food web structure from predator–prey body size relationships.
  *Methods in Ecology and Evolution*, 4, 1083–1090.
* Albouy, C. et al. (2019). The marine fish food web is globally connected.
  *Nature Ecology & Evolution*, 3, 1153–1161.
* Froese, R. & Pauly, D. (eds.). FishBase. World Wide Web electronic publication, www.fishbase.org.
* Rabosky, D.L. et al. (2018). An inverse latitudinal gradient in speciation
  rate for marine fishes. *Nature*, 559, 392–395.
* Stekhoven, D.J. & Bühlmann, P. (2012). MissForest: non-parametric missing
  value imputation for mixed-type data. *Bioinformatics*, 28, 112–118.
* OBIS. Ocean Biodiversity Information System. Intergovernmental Oceanographic
  Commission of UNESCO. www.obis.org.
