---
title: "Building a fish metaweb and local food-web indices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building a fish metaweb and local food-web indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 90
)
```

```{r load}
devtools::load_all()
library(metawebr)
library(ggplot2)
```

## Overview

`metawebr` infers a regional fish food web (the *metaweb*) from body size and
simple ecological traits. It then extracts the local food web of each site
(for example an eDNA sample or a survey area), computes network indices, and
can compare them with null models.

| Step | What happens | Function(s) |
|------|--------------|-------------|
| 1 | Explore predator–prey body-size records | `df_interaction_fish` |
| 2 | Calibrate the body-size niche model | `metaweb_mod_parameters()`, `get_fig_eval()` |
| 3 | Gather and clean traits for your taxa | `get_traits()`, `infer_traits()`, `clean_traits()` |
| 4 | Predict interaction probabilities for every pair of taxa | `apply_model_metaweb()` |
| 5 | Apply ecological corrections | `correct_metaweb_fish()` |
| 6 | Pick a probability threshold and binarise the metaweb | `get_proba()`, `correct_metaweb_proba()` |
| 7 | Build site × taxon matrices and compute local indices | `make_species_station_matrix()`, `get_indic_cells()` |
| 8 | Compare local indices with null models | `get_indic_null()`, `plot_indic_null()` |
| 9 | Plot food webs | `plot_tree_network()`, `plot_network_focus_sp()`, `plot_network_interactive()` |

The interaction model is the allometric niche model of Gravel et al. (2013).
A predator of body length $L_{pred}$ feeds on prey whose length falls within a
Gaussian niche. The niche optimum $o$ and width $r$ both scale with predator
size (all lengths on a log10 scale):

$$
o = a_0 + a_1 \log_{10} L_{pred}, \qquad
r = b_0 + b_1 \log_{10} L_{pred}, \qquad
P(\text{link}) = \exp\left(-\frac{(o - \log_{10} L_{prey})^2}{2r^2}\right)
$$

## 1. Predator–prey body-size data

The package ships `df_interaction_fish`, a compilation of `r nrow(df_interaction_fish)`
observed feeding events. Each row gives a predator, a prey and their lengths (cm).
Only four columns are used for calibration:

```{r data}
data("df_interaction_fish")
head(df_interaction_fish[, c("predator", "prey",
                             "standardised_predator_length", "si_prey_length")])
```

On a log scale, prey size increases with predator size, and the spread of prey
sizes widens too. The niche model captures exactly this pattern:

```{r data-plot, fig.height = 4.5}
ggplot(df_interaction_fish,
       aes(standardised_predator_length, si_prey_length)) +
  geom_point(alpha = 0.15, size = 0.8) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Predator length (cm, log scale)",
       y = "Prey length (cm, log scale)") +
  theme_minimal()
```

## 2. Calibrate the niche model

`metaweb_mod_parameters()` removes duplicate records and caps each
predator–prey pair at `observation_resample` observations, so that heavily
sampled pairs don't dominate the fit. It then estimates $a_0, a_1, b_0, b_1$ by
maximum likelihood, using simulated annealing (`GenSA`). With
`data_path = NULL` it uses `df_interaction_fish`; you can also pass a data
frame or the path of a CSV with the same columns.

Calibration takes about two minutes and involves random down-sampling, so it is
not run here:

```{r calibrate, eval = FALSE}
set.seed(1)
params <- metaweb_mod_parameters(path_par_output = "params_metaweb.txt")
params$calibrated_data
```

The other functions accept the parameters in three forms: the list returned
by `metaweb_mod_parameters()`, the file it wrote, or a plain numeric vector.
This vignette uses the values estimated from `df_interaction_fish`:

```{r pars}
pars <- c(a0 = -0.5616698, a1 = 0.9611534, b0 = 0.1247726, b1 = 0.1359744)
```

`get_fig_eval()` checks the fit for one predator. The left panel compares the
sizes of that predator's observed prey with all prey sizes and with the
predicted niche. The right panel shows the predicted link probability across
the whole predator–prey size space, overlaid with the observations.

```{r fit-eval, fig.width = 9, fig.height = 4.5}
get_fig_eval(data = df_interaction_fish, target_sp = "Gadus morhua", pars = pars)
```

## 3. Example communities and their traits

`df_presence_obis` is an example community dataset built from OBIS occurrence
records since 2000. It holds the 40 most recorded fish species at each of six
sites:

* three Mediterranean rocky reefs: Marseille, Medes Islands, Port-Cros;
* three tropical western Atlantic coral reefs: Florida Keys, Belize Barrier
  Reef, Bonaire.

```{r presence}
data("df_presence_obis")
head(df_presence_obis)
table(df_presence_obis$region)
```

The model needs one row per taxon, with **row names set to the taxon names**,
and these columns:

* `CommonLengthEstim`: common length (cm). If FishBase has no common length,
  it is estimated as 0.6 × maximum length.
* `TrophicLevel`: the FishBase trophic level estimate. It is used by the
  corrections and to choose the threshold.
* `DemersPelag`: position in the water column (FishBase categories).

`get_traits()` downloads these traits from FishBase through `rfishbase`.
Taxon names must use underscores (`Coris_julis`). For eDNA taxa assigned only
to a genus or a family, `infer_traits()` averages the traits of all FishBase
species in that group.

```{r traits-fishbase, eval = FALSE}
df_traits_obis <- get_traits(df_presence_obis, column_species = "species")
```

The result is stored as `df_traits_obis`, so this vignette runs offline.
`clean_traits()` reports missing values and fills them from other species of
the same genus. By default it then removes taxa that still have missing traits.
Here *Equetus punctatus* is not in FishBase and has no congener in the data, so
it is removed:

```{r traits}
data("df_traits_obis")
traits <- clean_traits(df_traits_obis)
```

`clean_traits()` only borrows values from congeners in your own table. A more
thorough option is `impute_traits_phylo()`. It predicts missing values with
random forests (missForest), using each species' position in the fish
phylogeny as extra information. It repeats this over the 100 complete fish
trees of Rabosky et al. (2018) and averages the results. It downloads the trees
and takes a few minutes, so it is not run here:

```{r traits-impute, eval = FALSE}
set.seed(1)
traits_imputed <- impute_traits_phylo(df_traits_obis, mc.cores = 4)
traits_imputed[!is.na(traits_imputed$imputed), ]

# Or in one step from taxon names: FishBase lookup, then imputation
traits_obis <- infer_traits(unique(df_presence_obis$species), method = "100matrix",
                            mc.cores = 4)
```

The fishtree trees cover ray-finned fishes only. To impute sharks and rays,
pass your own trees with `trees =`.

## 4. Predict interaction probabilities

`apply_model_metaweb()` applies the calibrated model to every pair of taxa.
In the resulting square matrix, **rows are prey and columns are predators**:
`mw["Chromis_chromis", "Muraena_helena"]` is the probability that the moray
eel eats the damselfish.

```{r metaweb}
mw <- apply_model_metaweb(traits, pars)
dim(mw)
round(mw[c("Chromis_chromis", "Coris_julis", "Diplodus_vulgaris"),
         c("Muraena_helena", "Scorpaena_scrofa", "Coris_julis")], 2)
```

## 5. Ecological corrections

Size alone predicts some implausible links. `correct_metaweb_fish()` applies
five corrections, each of which can be switched off:

* `herbivory`: herbivores have no fish prey. By default these are taxa with a
  trophic level < 2.4; on these reefs, mostly parrotfishes, surgeonfishes and
  damselfishes.
* `water_pos`: removes links between taxa that live in incompatible parts of the
  water column (for example bathydemersal vs. pelagic-oceanic). It also removes
  cannibalism.
* `small_fish`: taxa shorter than 10 cm have no fish prey.
* `PP_SP_add`: adds two basal nodes, `PrimaryProducer` and `SecondaryProducer`,
  as food for herbivores, planktivores and small fish.
* `exception_BigFish`: large (> 100 cm) low-trophic-level taxa feed on
  secondary producers only.

```{r correct}
mw_corr <- correct_metaweb_fish(data_meta = mw, data_traits = traits)
dim(mw_corr)
```

The first message lists the taxa treated as herbivores. FishBase trophic levels
can misclassify some taxa, so check this list. To change it, use
`set_herbivores()`. It stores the classification in a `Herbivore` column of the
trait table, which `correct_metaweb_fish()` then uses instead of the trophic
level. For example, to stop treating the scrawled cowfish as a herbivore:

```{r herbivores}
traits_edited <- set_herbivores(traits, remove = "Acanthostracion_polygonium")
sum(traits_edited$Herbivore)
```

You would then run `correct_metaweb_fish(mw, traits_edited)` again. The rest of
this vignette keeps the default classification.

### Checking the corrections

`compare_metawebs()` lists what the corrections changed, so you can check it by
hand. Each species pair appears once, with both directions of the interaction:

* `taxon_1`, `taxon_2`: the two taxa;
* `t1_eats_t2_before`, `t1_eats_t2_after`, `t1_eats_t2_correction`: the
  probability that taxon 1 eats taxon 2 before and after correction, and the
  correction that changed it (`NA` if unchanged);
* the same three `t2_eats_t1_*` columns for the other direction (`NA` for
  cannibalism, where both taxa are the same);
* the traits of both taxa (`t1_*`, `t2_*`).

Pairs are sorted by the highest probability, before correction, among their
changed links, so the most likely links that were removed come first. Changed
links below 0.01 both before and after are hidden (`min_proba = 0` shows them
all).

```{r corrections-pairs}
changes <- compare_metawebs(mw, mw_corr, traits)
nrow(changes)
table(c(changes$t1_eats_t2_correction, changes$t2_eats_t1_correction))

shown <- head(changes[, 1:8])
shown[, c(3, 4, 6, 7)] <- signif(shown[, c(3, 4, 6, 7)], 2)
shown
```

Use `level = "link"` instead for one row per direction.

With `level = "taxon"` you get one row per taxon, with how much each
correction changed its prey. For a probability metaweb these are expected
numbers of prey:

```{r corrections-taxa}
by_taxon <- compare_metawebs(mw, mw_corr, traits, level = "taxon")
cols <- c("prey_before", "prey_after", grep("^prey_change_", names(by_taxon), value = TRUE))
cbind(taxon = by_taxon$taxon, round(by_taxon[, cols], 1))[
  by_taxon$taxon %in% c("Muraena_helena", "Sarpa_salpa", "Tripterygion_delaisi"), ]
```

To review the changes in a spreadsheet, save the table with
`write.csv(changes, "metaweb_corrections.csv", row.names = FALSE)`.
`compare_metawebs()` works on any two metawebs, e.g. before and after
binarisation (`compare_metawebs(mw_corr, mw_bin)`); changes that were not made
by `correct_metaweb_fish()` have no `correction`.

## 6. Choose a threshold and binarise

Network indices need presence/absence links, so the probabilities must be
thresholded. `get_proba()` tries thresholds from 0.1 to 0.995. For each one it
computes the trophic levels implied by the binary web and sums their absolute
differences from the FishBase trophic levels. It keeps the threshold with the
smallest error:

```{r threshold, fig.height = 4.5}
threshold <- get_proba(mw_corr, traits)
threshold
```

`correct_metaweb_proba()` runs the same search and returns the binary metaweb.
You can also give it a `threshold` directly, for example to test how sensitive
your results are to this choice.

```{r binarise, fig.show = "hide"}
mw_bin <- correct_metaweb_proba(mw_corr, traits)
sum(mw_bin)  # number of links
```

## 7. Local food webs and indices

Local food webs are sub-networks of the metaweb: a link is kept when both taxa
occur at the site. `make_species_station_matrix()` turns long detection data
(one row per site × taxon, with a count such as eDNA reads or records) into a
binary site × taxon matrix. It also adds the two producer nodes to every site.

```{r stations}
pa <- make_species_station_matrix(df_presence_obis,
                                  station_col = "station",
                                  species_col = "species",
                                  count_col = "count")
dim(pa)
```

`get_indic_cells()` drops taxa that are absent from the metaweb, with a
warning. Here that is the species removed by `clean_traits()`, so we drop it
explicitly:

```{r stations-filter}
pa <- pa[, colnames(pa) %in% rownames(mw_bin)]
```

`get_indic_cells()` returns one row per site with about 40 indices. Its
`mc.cores` argument runs sites in parallel (not supported on Windows).

```{r indices}
indices <- get_indic_cells(P_A_data = pa, Lniche = mw_bin)

cols <- c("Species", "Connectance", "TL_moy", "length_chain", "Omn_moy",
          "Ntop", "Nbas", "Modularity")
cbind(site = indices$site, round(indices[, cols], 3))
```

Some of the indices:

| Index | Meaning |
|-------|---------|
| `Species` | Number of nodes, including the two producer nodes |
| `Link`, `Connectance` | Number of links, and links / nodes² |
| `TL_moy`, `length_chain` | Mean trophic level, and maximum trophic level rounded up |
| `Omn_moy` | Mean omnivory index |
| `Ntop`, `Nbas`, `Nint` | Fractions of top (no predator), basal (at most one prey) and intermediate taxa |
| `Vul`, `Gen` | Mean number of predators and of prey per taxon |
| `Modularity` | Modularity of walktrap communities (`igraph`) |
| `Mean_path_length`, `Shortest_path_k` | Mean shortest path, and the share of paths of length *k* |
| `Planktivores`, `HTI`, `MTI` | Share of planktivores, high-trophic-level indicator and mean trophic level of predators. Computed only for webs with more than 50 nodes, otherwise `NA` |

See `?get_indic_cells` for the full list.

## 8. Null models

Many indices change with the number of species, so raw values are hard to
compare between sites. A null model asks whether a local web is more or less
connected, modular, etc. than expected for a random community of the same
richness, drawn from the regional pool.

`get_indic_null()` builds `n_null` random communities per site, keeping the
site's richness and its producer nodes, and computes the same indices. For each
index it reports:

* the null mean and standard deviation;
* the standardised effect size, `ses = (observed − null_mean) / null_sd`.
  Values below −1.96 or above 1.96 are roughly significant at 5 %;
* a two-sided, rank-based `p_value`.

Two methods are available:

* `"equiprobable"`: every taxon of the pool is equally likely to be drawn;
* `"frequency"`: taxa are drawn in proportion to the number of sites where they
  occur, so common taxa appear more often.

The pool is every taxon present in at least one site of `P_A_data`. Pooling the
Mediterranean and Caribbean sites would put Caribbean species into
Mediterranean null communities, so we run the null model separately for each
region:

```{r null-equiprobable}
region <- setNames(df_presence_obis$region, df_presence_obis$station)
sites_by_region <- split(rownames(pa), region[rownames(pa)])

set.seed(42)
null_equi <- do.call(rbind, lapply(sites_by_region, function(sites) {
  get_indic_null(pa[sites, ], mw_bin, method = "equiprobable", n_null = 199)
}))
rownames(null_equi) <- NULL

res <- null_equi[null_equi$index %in% c("Connectance", "TL_moy"), ]
cbind(res[, c("site", "index")], round(res[, c("observed", "null_mean", "ses", "p_value")], 3))
```

`plot_indic_null()` shows the effect sizes of several indices at once. Points
are coloured when p < 0.05 and dashed lines mark SES = ±1.96:

```{r null-plot, fig.width = 8, fig.height = 5.5}
plot_indic_null(null_equi,
                indices = c("Connectance", "TL_moy", "Omn_moy",
                            "Modularity", "Mean_path_length", "Nbas"))
```

`get_indic_cells()` can compute the observed indices and the null model in one
call. With `null_model = "none"` (the default) it returns the indices only;
with a method it returns a list with `indices` and `null_model`. Here is the
frequency-weighted model for the Mediterranean sites, compared with the
equiprobable one:

```{r null-frequency}
set.seed(42)
med <- get_indic_cells(pa[sites_by_region$Mediterranean, ], mw_bin,
                       null_model = "frequency", n_null = 199)
names(med)

compare <- merge(null_equi[, c("site", "index", "ses")],
                 med$null_model[, c("site", "index", "ses")],
                 by = c("site", "index"), suffixes = c("_equiprobable", "_frequency"))
compare[, c("ses_equiprobable", "ses_frequency")] <-
  round(compare[, c("ses_equiprobable", "ses_frequency")], 2)
compare[compare$index %in% c("Connectance", "TL_moy"), ]
```

`n_null = 199` keeps this vignette fast. Use 999 or more for published analyses.

## 9. Plotting food webs

`plot_tree_network()` draws a food web with each node's height set by its
trophic level. Nodes are coloured by rounded trophic level and sized by number
of prey. Horizontal positions are random, so set a seed for a reproducible
layout. The result is a `ggplot`, so you can add titles and themes as usual.
When nodes are crowded, some labels are left out.

Use `sp_to_keep` to plot a local web. Trophic levels are still computed from the
full metaweb, so node heights are comparable between sites:

```{r plot-local, fig.width = 9, fig.height = 7, warning = FALSE}
marseille <- df_presence_obis$species[df_presence_obis$station == "Marseille"]
set.seed(1)
plot_tree_network(mw_bin, sp_to_keep = marseille) + ggtitle("Marseille")
```

`plot_network_focus_sp()` zooms in on one or more focal taxa and their prey, up
to `order_interaction` steps down the food chain (`2` adds the prey of the prey):

```{r plot-focus, fig.width = 8, fig.height = 6, warning = FALSE}
set.seed(1)
plot_network_focus_sp(mw_bin, sp_focus = "Muraena_helena",
                      order_interaction = 2, sp_to_keep = marseille) +
  ggtitle("Moray eel and its prey in Marseille")
```

`plot_network_interactive()` draws the same layout as an interactive network
(it needs the visNetwork package). Click a taxon to highlight its links and
fade the rest of the web; hold Ctrl (Cmd on macOS) while clicking to select
several taxa, or pick one by name in the menu. Click an empty area to clear
the selection. Hover a taxon or a link for details, scroll to zoom and drag to
pan. With `links = "prey"` or `links = "predators"` only one direction is
highlighted:

```{r plot-interactive, eval = requireNamespace("visNetwork", quietly = TRUE)}
set.seed(1)
plot_network_interactive(mw_bin, sp_to_keep = marseille, height = "600px")
```

The result is an htmlwidget, so you can save it as a stand-alone HTML file to
share:

```{r save-interactive, eval = FALSE}
web <- plot_network_interactive(mw_bin, sp_to_keep = marseille)
htmlwidgets::saveWidget(web, "marseille_foodweb.html")
```

## References

* Gravel, D., Poisot, T., Albouy, C., Velez, L. & Mouillot, D. (2013). Inferring
  food web structure from predator–prey body size relationships.
  *Methods in Ecology and Evolution*, 4, 1083–1090.
* Albouy, C. et al. (2019). The marine fish food web is globally connected.
  *Nature Ecology & Evolution*, 3, 1153–1161.
* Froese, R. & Pauly, D. (eds.). FishBase. World Wide Web electronic publication, www.fishbase.org.
* Rabosky, D.L. et al. (2018). An inverse latitudinal gradient in speciation
  rate for marine fishes. *Nature*, 559, 392–395.
* Stekhoven, D.J. & Bühlmann, P. (2012). MissForest: non-parametric missing
  value imputation for mixed-type data. *Bioinformatics*, 28, 112–118.
* OBIS. Ocean Biodiversity Information System. Intergovernmental Oceanographic
  Commission of UNESCO. www.obis.org.
